---
title: "Graticule technical exercise"
author: "Ariel Asper"
date: "12/3/2023"
output: html_document
---

<ul>libraries used in analysis:
<li>dplyr
<li>lubridate
<li>data.table
</ul>

```{r,echo=FALSE,warning=FALSE}
#Libraries used:
library(dplyr)
library(lubridate)
library(data.table)

```

## 1. Load the data tables into dataframe(s) or a database
```{r}
patient_table <- read.csv('C:/Users/Ariel/Downloads/patient_table.csv')
encounter_table <- read.csv('C:/Users/Ariel/Downloads/encounter_table.csv')
```


## 2. Compute median age at index date for the exposed cohort (including CI at 95% confidence level)

### Frist, lets convert the date variables into type "Date" variables and remove all encounters not in study period:
```{r}
patient_table$birthdate <- as.Date(patient_table$birthdate)
patient_table$deceaseddate <- as.Date(patient_table$deceaseddate)
encounter_table <- encounter_table%>%
                   mutate(start_date = as.Date('2022-07-01'),#As stated in guidelines
                          end_date= as.Date('2022-12-31'),#As stated in guidelines
                          encounterdate = as.Date(encounterdate))%>%
  filter(encounterdate>=start_date)#filter out all encounters not in study period ie., keep only those with encounter dates from July 1st though Dec 31st 2022

```

### Next, we will identify our exposure cohort and calcualte median(including CI at 95% confidence level)

```{r,warning=FALSE}
# lets see who will enter exposure group. This will be done by filtering out those who were never exposed.
# In cases where individual was exposed multiple times,index date will be date at first exposure.
Exposed <- encounter_table%>%
  filter(exposure=='True')%>%#Extract only those who have been exposed at some point during study period.
  group_by(patientid)%>%
  arrange(encounterdate)%>%
  filter(row_number()==1)%>%#extract first encounter where exposed
  rename("index_date"='encounterdate')#first date when exposed is identified as index date.

#Lets merge Exposed df with patient in order to get birth date data 
Exposed <- merge(Exposed,patient_table,by=c('patientid'),all.x = T)

#calculate age at index date for exposed cohort
Exposed <- Exposed%>%
  mutate(Index_date_age=round(as.numeric((index_date-birthdate)/365)))%>%
  select(-exposure,-outcome,-start_date,-end_date)#remove unnecessary variables 

#Calculating median of age at Index date
med = median(Exposed$Index_date_age)
#Calculating CI at 95% confidence level using Wilcoxon test.
wt <- wilcox.test(Exposed$Index_date_age,conf.int = T)

#Print results
print(paste0('Median age at index date for the exposed cohort is:',med, ' with CI of [',round(wt$conf.int[1]),'-',round(wt$conf.int[2]),']',' at 95% confidence level'))
```

#### <b>Answer:</b> Median age at index date for the exposed cohort is:50 with CI of [45-60] at 95% confidence level


## 3. Compute aggregate time at risk for the exposed cohort, in person-days

<ol><b>In order to do so we need to:</b>
<li> Define end date for follow-up period.
<li> Calculate, for each individual, days until end of follow-up.
<li> The Sum of all days at risk for exposure group will give us the time at risk in person days.

End of follow-up will be Dec 31st 2022 or death. Whichever comes first.

```{r}
df <- merge(Exposed,encounter_table,by=c('patientid'),all.x = T)
#first event date
df <- df %>% group_by(patientid)%>%
  mutate(event_date=fifelse(outcome=="True",encounterdate,as.Date(NA)))%>%
  mutate(event_date = fifelse(event_date<index_date,as.Date(NA),event_date))

df$end_date <- pmin(as.Date('2022-12-31'),df$deceaseddate,df$event_date,na.rm = T)

#Calculate days at risk
PD = df%>%
  group_by(patientid)%>%
  summarise(end_date = min(end_date),index_date = min(index_date))%>%
  mutate(at_risk_days = 1+(end_date-index_date))%>%summarise(sum_person_days=sum(at_risk_days))
print(paste0('time at risk for the exposed cohort, in person-days is: ',as.numeric(PD),' person days'))
```


#### <b>Answer:</b> time at risk for the exposed cohort, in person-days is: 1820 person days



##  Finally, provide pseudo-code to select a 1:1 unexposed cohort based on the method outlined above.


#### Input:
  
<ul>
<li>full_df: Dataframe containing patient information in long format. Each row detailing encounter information and demographic information.
<li>match_vars: Vector of variables for exact matching.



#### Output:
<ul><li>matched_data: Dataframe containing matched pairs of exposed and control units.</ul>

### Algorithm:
  
#### Data Preparation:

<ol>
<li>Initialize empty DataFrame called "matched_data".
<li>Set group indicator to 1.
<li>Filter out all encounter dates after index date for those exposed.
<li>Add a new variable : is_control (logical).
<li>Initialize is_control to FALSE for exposed individuals and TRUE for controls.

<li>for each encounter in DataFrame:</ul>
|       If is_control is FALSE:
|           If yes:
|             Check if encounter was at least 7 days prior to index_date: 
|             If yes, set is_control to TRUE for that encounter
|           If no:
|             Delete that row
|


<li>Split the DataFrame into two DataFrames
<ul><li>Exposure: Those with is_control set to FALSE
<li>No exposure:Those with is_control set to TRUE
<li> Create copies of both DataFrames</ul>


#### Matching loop:


| For each individual in the exposed DataFrame:
|   a. For each individual in the no_exposure DataFrame:
|       i. Check if the no_exposure individual matches the exposed individual on match_vars.
|       ii. If yes:
|           a. Check if the no_exposure individual's encounter date is within 7 days of the exposed individual's index_date.
|           b. If yes:
|               i. Check if the no_exposure individual is already in matched_data with is_control set to TRUE.
|               ii. If no:
|                   1. Add the pair to matched_data with the current group_indicator.
|                   2. Remove both individuals from their respective DataFrames.
|                   3. Break and return to the first loop.
|   b. Increment group_indicator by one.


#### Handle unmatched units:
  
<ul>
<li>If there are remaining exposed units with is_control = FALSE, they are not matched. Consider handling them separately (e.g., exclude from analysis).
<li>If there are remaining control units with is_control = TRUE, they were not used as matches. Consider handling them separately (e.g., exclude from analysis or use as additional controls).
<li> Other things to conciser in this situation could be:
<ul> 
<li> Selecting other matching method.
<li> Matching on less variables.
<li> Allowing for some differnce between cases and controls(e.g., allowing to be matched if older or younger by X years).
</ul></ul>
####   Combine results:
  
Join matched_data with the original exposed and control DataFrames on common identifiers (e.g., ID variable) to retain additional information.

### Return:
  
   <b>matched_data</b>


In practice, using matchit function from Matchit library would be more efficient.

